### 2017-11-20 manuscript.rmd
removed floating section header, added run time to sup info
revised implications section

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree. The other reason is that pairwise comparisons and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. Any difference could be due to the structure of the gene phylogenies alone. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

@@

@@

There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi; @Altenhoff:2012ea]. While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk]. Since the ortholog conjecture can be applied to many different gene attributes at both the genotype level (*e.g.*, molecular sequence) and phenotype level (*e.g.*, activity, biochemical function, binding partners, and expression), it is naturally a heterogeneous and potentially idiosyncratic topic [@Soria:2014bv]. For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression than expected [@Soria:2014bv].

Our results suggest, at a minimum, that the ortholog conjecture is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This suggests that an alternative "neutral conjecture", *i.e.* the conjecture that the evolution of gene traits tends to be the same following gene duplication and speciation events, may better explain the process and patterns of most gene evolution. Under this neutral conjecture, phylogenetic distance is a better predictor of the similarity of gene function than is the history of gene duplication and speciation. The ortholog conjecture does not have to be an all or nothing process, though. It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi]. We just do not find evidence for it when summarizing gene expression across tissues with tau in these organisms. This calls into question the general predictive power of the ortholog conjecture with respect to gene expression, and until these processes are better understood it will be necessary to test for it in each situation. These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.

Lack of support for the ortholog conjecture has important biological implications. It indicates that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed. It also has important technical implications. Having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].

This is also an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the phylogeny that are derived from the structure of the tree and history of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.

@@

#### Origin of pairwise correlation plot structure

@@

These analyses were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.

>>>>

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree. The other reason is that pairwise comparisons and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. Paralogs can be more different than orthologs even when the processes of evolution following speciation and duplication are the same. Any difference could be due to the structure of the gene phylogenies alone. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance (*i.e.*, total branch length) between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

@@

	edge_summary_pass_k = edge_summary %>% filter( gene %in% genes_pass_k )

	edge_pass_k_p = wilcox.test( 
		edge_summary_pass_k %>% filter( event_parent=="Duplication" ) %>% .$scaled_change, 
		edge_summary_pass_k %>% filter( event_parent=="Speciation" ) %>% .$scaled_change, 
		alternative="greater" 
	)$p.value

@@

There has been considerable recent interest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi; @Altenhoff:2012ea]. While some studies have presented support for the ortholog conjecture, our results are consistent with multiple studies that have not [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk]. It should be noted that many areas of biology restrict the term "function" to describe the biochemical function of a gene, but the literature on gene duplication often uses "function" in a broader sense to include any gene attribute that could impact fitness, such as biochemical function, spatial expression, or binding. Since the ortholog conjecture can be applied to such diverse attributes, it is a heterogeneous and potentially idiosyncratic topic. For example, a recent integrated analysis of multiple gene traits after duplication found extensive divergence in protein sequence but less divergence in expression [@Soria:2014bv]. In our analyses, we found that phylogenetic distance is a better predictor of the similarity of gene expression than is the history of gene duplication and speciation. The ortholog conjecture does not have to be an all or nothing proposition, even as it applies to expression. It may be that the rates of tissue-specific expression evolution following duplication are greater in some organisms, gene families, and evolutionary processes [@Gabaldon:2013fi]. The lack of support we find for the ortholog conjecture in this system, as well as the mixed results of others, does suggests that investigators should test for it in each situation rather than assume *a priori* that it is a dominant pattern in the diversity of gene expression. These tests should be done in an explicit phylogenetic framework.

While empirical support for the ortholog conjecture has been mixed, theoretical work has suggested several mechanisms that could lead to increased rates of evolutionary change in expression and other gene traits following gene duplication. This has been a primary motivation for the extensive discussion of the OC. In his seminal work, Ohno [@Ohno:1970wm] outlined three outcomes of gene duplication, now often referred to as neofunctionalization (one gene copy can take on new functions while the other retains the old functions), subfunctionalization (each gene copy has a subset of the functions present in their most recent common ancestor), and conservation (each gene copy retains the same function). Neofunctionalization was the primary focus of early work on gene duplication, though it was not found to be a widespread outcome of duplication [@Hughes:1994ec]. The highly influential duplication-degeneration-complementation (DDC) model [@Force:1999tg; @Lynch:2000um] suggested that subfunctionalization could be a better explanation for the retention of both copies of a gene. It is built on the idea that gene copies that have each lost different functions would both be needed to fulfill all the functions present in their most recent common ancestor. Tests of the DDC model have had mixed results [@Huminiecki:2004dc]. A growing body of work suggests that retention and differentiation of genes is due to a variety of different mechanisms in different contexts, and that we should not necessarily expect one process to dominate over others [@Hahn:2009by]. Our lack of support for the ortholog conjecture here suggests that whatever mechanisms lead to the retention of duplicated genes in the organisms considered here does not lead to changes in the tissue-specificity of gene expression as summarized by tau.

The DDC model is particularly interesting to further consider in this context since the summary statistic tau describes tissue specificity of expression. In the context of tissue-specific expression, the OC predicts a greater rate of evolutionary change in tau following duplication, without specifying the direction of the change. The DDC model predicts a specific type of change in expression following duplication -- subfunctionalization, which would increase tissue specificity, and therefore tau, following duplication. To test for this more specific predictions of DDC, we conducted additional analyses to see if there is a greater rate of increase in tau following duplication. For each branch in each gene tree, we inferred the change in tau and standardized this change by branch lengths. We found no evidence of increased tau following duplication relative to speciation (Wilcoxon $P=`r  edge_p %>% round( 3 )`$). This result also holds when we discarded the `r k_percentile` of genes with the lowest phylogenetic signal (Wilcoxon $P=`r  edge_pass_k_p %>% round( 3 )`$). These results are consistent with another recent study that found little support for neofunctionalization or subfunctionalization in expression of duplicated human genes, and instead proposed that dosage compensation drives most changes in expression following duplication in mammals [@Lan:2016ix]. Like the OC, DDC is not an all-or-nothing hypothesis. We fail to find support for it in these tissues in these species with the tau expression summary statistic, but it may hold in other contexts. DDC could still play a role in the retention of these genes, but with respect to other attributes such as biochemical functions rather than tissue-specific expression.

@@

@@

These analyses were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions. The total run time of these analyses was `r time_run %>% as.numeric() %>% round( 2 )` hours on a `r cores+1` core machine.
